\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2017-09-25 paper/casebase_jss.Rmd
add theoretical details

>>>>

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it. We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*). The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.

Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.

As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling. This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

### 2017-09-25 paper/casebase_jss.tex
add theoretical details

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.81 & (0.47, 1.41)\\
\hline
Disease & 0.57 & (0.31, 1.03)\\
\hline
Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.01)\\
\hline
Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.8)\\
\hline
Phase (Relapse vs. CR1) & 4.23 & (1.97, 9.09)\\
\hline
Source & 1.60 & (0.53, 4.88)\\
\hline
Age & 1.00 & (0.97, 1.02)\\
\hline
\end{tabular}

>>>>

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to discretize the study base into
an infinite amount of \emph{person moments}. These person moments are
indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure status
and an outcome status attached to it. We note that there is only a
finite number of person moments associated with the event of interest
(what Hanley \& Miettinen call the \emph{case series}). The case-base
sampling refers to the sampling from the base of a representative finite
sample called the \emph{base series}.

Popular survival analysis methods, like Kaplan-Meier and Cox regression,
rely on the notion of risk-set sampling. Case-base sampling can be seen
as an alternative.

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogenous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
the counting process associated with the event of interest,
\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
the hazard function for the Poisson process associated with case-base
sampling. This parametric form suggests that we can readily estimate
log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
regression, where each observation corresponds to a person moment, the
function \(g(t; X)\) is linear in a finite number of parameters, and
where we treat \(\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base samping \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds to a time-homogeneous
Poisson process with intensity equal to \(b/B\), where \(b\) is the
number of sampled observations in the base series, and \(B\) is the
total population-time for the study base. More complex examples are also
available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.82 & (0.46, 1.48)\\
\hline
Disease & 0.48 & (0.25, 0.92)\\
\hline
Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
\hline
Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
\hline
Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
\hline
Source & 1.51 & (0.49, 4.69)\\
\hline
Age & 0.99 & (0.97, 1.02)\\
\hline
\end{tabular}
